Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-06-17 paper/ms.tex
Several small text updates

  Elongated bar-like features are ubiquitous in galaxies, occurring at the
  centers of approximately two-thirds of spiral
  disks\cite{2000AJ....119..536E, 2007ApJ...657..790M}.  Due to gravitational
  interactions between the bar and the other components of galaxies, it is
  expected that angular momentum and matter will redistribute over long (Gyr)
  timescales in barred galaxies \cite{1972MNRAS.157....1L,
  1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work ignoring the gas
  phase of galaxies has conclusively demonstrated that bars should slow their
  rotation over time due to their interaction with dark matter halos
  \cite{1992ApJ...400...80H, 2000ApJ...543..704D, 2002MNRAS.330...35A,
  2002ApJ...569L..83A, 2003MNRAS.341.1179A, 2003MNRAS.346..251O,
  2005MNRAS.363..991H, 2006ApJ...637..214M, 2007MNRAS.375..460W,
  2009ApJ...697..293D}. We have performed a simulation of a Milky Way-like
  galactic disk hosting a strong bar which includes a state-of-the-art model
  of the interstellar medium and a live dark matter halo. In this simulation
  the bar pattern does not slow down over time, and instead remains at a
  stable, constant rate of rotation. This behavior has been observed in
  previous simulations using more simplified models for the interstellar gas
  but it has remained unexplained\cite{1993AA...268...65F,
  2010ApJ...719.1470V}. We propose that the gas phase of the disk and the dark
  matter halo act in concert to stabilize the bar pattern speed and prevent a
  bar from slowing down or speeding up. We find that in a Milky Way-like disk,
  a gas fraction of only about \boldmath$5\%$ is necessary for this mechanism
  to operate. This result naturally explains why nearly all observed bars
  rotate rapidly\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
  2019MNRAS.482.1733G} and is especially relevant for our understanding of how
  the Milky Way arrived at its present state.

@@

\begin{figure*}[h!]%
\centering
\includegraphics[width=18cm]{fig/halo_wake.pdf}
\caption{\textbf{The wake excited in the dark matter halo.} The dark matter
halo wake is shown in the \Nbody{} case (\textit{middle panel}) and SMUGGLE case
(\textit{right panel}) after $2.6\,\textrm{Gyr}$ of evolution. The
\textit{middle} and \textit{right panels} show a surface density projection in
the $x$-$y$ plane of the dark matter halo after an axisymmetric average has
been subracted. The solid line indicates the direction of the bar while the
dashed line indicates the direction of the halo wake (both measured by taking
the second Fourier component within a sphere of all material within a radius
of $4\,\textrm{kpc}$). The \textit{left panel} shows the time evolution of
the angle difference between the bar and the halo wake, as measured from the
second Fourier component. After the first Gyr, the angle difference in the
SMUGGLE case is smaller than in the \Nbody{} case by about a factor of two,
reflecting how the dark matter halo in the SMUGGLE case is unable to exert as
negative a torque on the bar as in the \Nbody{} case.}\label{fig:wake}
\end{figure*}

@@

It has long been known that non-axisymmetric features in a stellar disk act to
redistribute mass and angular momentum.\cite{1972MNRAS.157....1L} The
interaction between a bar and a spheroid (either a dark matter halo or stellar
bulge) has received considerable interest.\cite{1984MNRAS.209..729T,
1985MNRAS.213..451W} Stellar disks in isolation are prone to developing bars,
\cite{1971ApJ...168..343H} but the presence of a strong spherical potential
(e.g., from a stellar bulge or dark matter halo) acts to stabilize the disk
against bar formation.\cite{1973ApJ...186..467O, 1976AJ.....81...30H} Many
numerical simulations have confirmed that a live dark matter halo generically
exerts a negative torque on a bar, causing its pattern speed to decrease and
its length and strength to increase\cite{1992ApJ...400...80H,
2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. Intuitively,
the bar excites a wake of resonant material in a dark matter halo in a manner
akin to dynamical friction. This wake lags the bar and thus exerts a negative
torque on it. For a review of bar dynamics, see ref.\cite{1993RPPh...56..173S}

The expectation that a dark matter halo acts to slow down a bar is in
conflict with observational estimates of bar pattern speeds. Bar rotation
rates are typically classified by the dimensionless ratio
\begin{equation}
\Rot = \RCR/R_b\textrm{,}
\end{equation}
where \RCR{} is the radius of corotation and $R_b$ is the length of the
bar.\footnote{The radius of corotation \RCR{} is defined for circular orbits
as the radius at which the orbital frequency is equal to the pattern speed,
$\Omega_p$, of a given non-axisymmetric feature. In a galaxy with a constant
circular velocity $V_c$, it is given by $\RCR = V_c / \Omega_p$.} Galaxies
with $\Rot < 1.4$ are considered ``fast rotators'' while galaxies with $\Rot >
1.4$ are considered ``slow rotators.''\cite{2000ApJ...543..704D} Galaxies with
$\Rot < 1$ are not thought to be stable.\cite{1980AA....81..198C}
Observational estimates of the pattern speeds of bars indicate that nearly all
galaxies have $1 < \Rot < 1.4$.\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G, 2020MNRAS.491.3655G}

The role of gas on the evolution of the bar is less well-understood. Since the
gas phase typically only contributes about $10-20\%$ of the mass of a galaxy
at the present day, one might naively expect it to have a subdominant effect
on the bar. However, because gas is collisional, it can participate in
non-resonant angular momentum exchange with the bar.\cite{2011MNRAS.415.1027H}
Thus, numerical work has shown that the gas phase can have a stronger
influence on a bar than its contribution to the mass of a galaxy would
suggest.\cite{2010ApJ...719.1470V, 2013MNRAS.429.1949A}

We have performed a simulation of a disk galaxy using the finite-volume
gravito-hydrodynamics code \AREPO{}.\cite{2010MNRAS.401..791S} In \AREPO{},
the fluid is discretized as a moving Voronoi mesh. We use the additional
physics in the galaxy formation module Stars and MUltiphase Gas in GaLaxiEs
(SMUGGLE) \cite{2019MNRAS.489.4233M}. SMUGGLE is a comprehensive and
self-consistent galaxy formation model with a wide range of physical
processes, including radiative heating/cooling, star formation, and stellar
feedback. More detailed information on SMUGGLE is given in the Methods
section. Our disk galaxy is a modified version of the GALAKOS
model\cite{2020ApJ...890..117D}, which consists of a stellar disk, stellar
bulge, and dark matter halo. After $\sim2.5\,\textrm{Gyr}$ of evolution, the
GALAKOS disk forms a bar consistent with the Milky Way bar in terms of pattern
speed and length
($\sim40\,\textrm{km}/\textrm{s}/\textrm{kpc}$\cite{2022arXiv220606207G} and
$\sim4.5\,\textrm{kpc}$\cite{2015MNRAS.450.4050W}, respectively). We modify
this setup to include a gas phase, the details of which are provided in the
Methods section.

A face-on surface density projection of our simulations is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the \Nbody{}
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large
qualitative difference in the evolution of the bar pattern between the two
runs. We see that in the \Nbody{} case, the bar lengthens in time and grows in
strength. In the SMUGGLE case, the bar retains a similar length and strength
between panels.

We show the time evolution of different bar properties in Fig.~\ref{fig:prop}.
In the upper panel, we show the pattern speed over time in the \Nbody{} (blue)
and SMUGGLE (orange) runs. The pattern speed in the \Nbody{} case slows down
while the pattern speed in the SMUGGLE case remains roughly constant. The
slowing down of the pattern speed in the \Nbody{} case is consistent with a long
line of numerical research on bars in \Nbody{}
simulations.\cite{1992ApJ...400...80H, 2000ApJ...543..704D,
2002MNRAS.330...35A, 2002ApJ...569L..83A, 2003MNRAS.341.1179A,
2003MNRAS.346..251O, 2005MNRAS.363..991H, 2006ApJ...637..214M,
2007MNRAS.375..460W, 2009ApJ...697..293D} However, in the SMUGGLE case the
pattern speed remains constant. After the first Gyr of evolution, we find that
the pattern speed increases by only $\sim10\%$ over the next
$4\,\textrm{Gyr}$, compared to a $\sim43\%$ decrease in the pattern speed for
the \Nbody{} run over the same interval. As we saw qualitatively in
Fig.~\ref{fig:overview}, the length of the bar in the \Nbody{} case grows over
time while it remains roughly constant in the SMUGGLE case. This is also
consistent with previous numerical work, which found that bars tend to grow as
they slow down and the radius of corotation
increases.\cite{2000ApJ...543..704D, 2003MNRAS.341.1179A}

@@

The fact that the dark matter halo in the SMUGGLE case exerts a smaller
negative torque on the bar can be understood in terms of the halo wake
mechanism. In the \Nbody{} case, halo material which is resonant with the bar
will form a wake which lags behind and exerts a negative torque on the bar,
which slows it down.\cite{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its
pattern speed. However, the bar seems to empirically respond to a negative
torque induced by a halo wake by slowing down.} As the bar slows down, the
location of the resonances in phase space changes, allowing halo material
newly resonant with the bar to participate in forming the wake. However, the
gas is a reliable source of positive torque on the bar, speeding the bar up.
This stops the resonance location from changing such that the halo cannot
reinforce a wake, arresting the process by which the halo can slow the bar
down. We term this process ``exhaustion.''

@@

The presence of the gas can arrest the process by which additional material in
the dark matter halo can contribute to a wake. However, this does not explain
why the pattern speed in the SMUGGLE case is nearly constant over several Gyr.
Naively, it would be a coincidence that the bar pattern speed remains constant
in the SMUGGLE case, resulting from a chance cancellation of the halo and gas
torques. However, a constant pattern speed in the presence of gas has been
observed in a few simulations of barred galaxies with
gas.\cite{1993AA...268...65F, 2010ApJ...719.1470V} Previous work has argued
this is due to the bar torquing gas inwards, but no explanation has been given
for why it might remain constant.

We propose that an equilibrium mechanism is responsible for the pattern speed
remaining approximately constant. In this scenario, residual negative torque
from the dark matter halo balances out the positive torque from the gas phase.
It has been shown when an analytic bar is forced to rotate at a constant
pattern speed for a few Gyr, the halo exerts almost no torque on the
bar.\cite{2022MNRAS.513..768C} We instead see over a several Gyr time span
that the dark matter halo in our simulation is still able to support some
negative torque.

@@

We note that ref.\cite{2020MNRAS.491.3655G} found that the rotation parameter
\Rot{} positively correlates with gas fraction, such that galaxies with higher
gas fractions are rotating more slowly. However, it is not obvious this is in
tension with our result since the gas fraction of galaxies correlates with
other galactic properties.\cite{2009ARAA..47..159B} Furthermore, the
measurement of pattern speeds is a delicate process still prone to large error
bars.

The implications of our findings are numerous. First, we naturally explain why
nearly all observed galaxies are fast rotators without requiring the inner
regions of dark matter halos to be underdense\cite{1998ApJ...493L...5D,
2000ApJ...543..704D} or requiring new physics.\cite{2021MNRAS.503.2833R,
2021MNRAS.508..926R} Second, we show that the role of gas is of paramount
importance in studies which attempt to uncover the nature of dark matter from
its effect of slowing down the bar.\cite{2021MNRAS.500.4710C,
2021MNRAS.505.2412C} Third, we provide an explanation for how the Milky Way's
bar could be both long-lived and a fast rotator, of which there is some
observational evidence.\cite{2019MNRAS.490.4740B} And finally, we complicate
the picture of radial mixing expected to sculpt the Milky Way's disk
\cite{2012MNRAS.420..913B, 2015ApJ...808..132H}, a process which relies upon
the pattern speed of the bar to change with time (though our work does not
alter expectations for radial mixing induced by spiral
arms\cite{2002MNRAS.336..785S}).

Barred galaxies in cosmological simulations of galaxy formation continue to be
in conflict with observations by producing bars which rotate too
slowly.\cite{2017MNRAS.469.1054A, 2019MNRAS.483.2721P, 2021AA...650L..16F}
However, the pattern speeds of bars in both cosmological simulations and the
real universe can be affected by environmental processes not included in our
simulation -- e.g., satellite infall\cite{2011Natur.477..301P},
non-sphericity\cite{2013MNRAS.429.1949A} or rotation\cite{2013MNRAS.434.1287S,
2014ApJ...783L..18L, 2018MNRAS.476.1331C, 2019MNRAS.488.5788C} in the dark
matter halo, or perhaps even the gaseous circumgalactic medium. Naturally,
extending our present work to account for such effects is a crucial next step
in understanding the formation and evolution of galactic bars.

We found that below a certain gas fraction, bars should still be able to slow
down. Therefore, we expect barred spiral galaxies which have been gas-poor for
extended periods of time to be rotating very slowly. We therefore predict that
observations which target such galaxies (e.g., lenticular barred
galaxies\cite{2009ARAA..47..159B}) would find slowly rotating bars. There does
exist one example of a galaxy which is known to be a slow rotator -- the
low surface brightness galaxy UGC 628.\cite{2009AA...499L..25C} This galaxy
has been studied in detail by ref.\cite{2016MNRAS.463.1751C}, who note that it
indeed has a low gas fraction for galaxies of its type. We predict a general
trend that bars in gas-rich spiral galaxies should rotate quickly while bars
in gas-poor spiral galaxies should rotate slowly.

@@

\item[Data Availability]  Simulation output and initial conditions will be
made publicly available.

\item[Code Availability] A public version of the \AREPO{} code is available at
\url{https://arepo-code.org/}. An associated GitHub repository containing
analysis and plotting scripts will be made publicly available (currently
private). For the submission stage, we have included a tarball of this
repository.

@@

\noindent
{\bf Initial setup}
\\
\noindent
The initial setup of the galactic disk used in this work follows closely the
GALAKOS model\cite{2020ApJ...890..117D}, which uses a modified version of
\texttt{MakeNewDisk}.\cite{2005MNRAS.361..776S} The GALAKOS model has three
components - a radially exponential and vertically isothermal stellar disk,
and a stellar bulge and dark matter halo following a Hernquist
profile.\cite{1990ApJ...356..359H} All \Nbody{} runs in this work used the same
setup parameters as the GALAKOS disk, more details of which can be found in
the original paper.

The addition of the gas phase was done as follows. The version of
\texttt{MakeNewDisk} used for the original GALAKOS model can generate a gas
disk which is radially exponential and in vertical gravito-hydrodynamic
balance. We modified the radial profile of this code in order to allow us to
generate a disk with a constant surface density within some cut-off radius,
and then exponentially declining beyond that radius with the scale-length of
the stellar disk. We used an initial surface density of
$20\,M_{\odot}/\textrm{pc}^2$ and a cut-off radius of $9.3\,\textrm{kpc}$. The
initial gas disk is generated with a temperature of $10^4\,\textrm{K}$ and
solar metallicity.

After generating the gaseous disk in this way, we stitched the gas disk
together with the GALAKOS \Nbody{} disk (and bulge and dark matter halo) after
the GALAKOS disk has been allowed to evolve for $1.5\,\textrm{Gyr}$. The
purpose of allowing the GALAKOS disk to evolve first for a short period of
time is to allow for the bar to form unimpacted by the presence of the gas
(which would normally disrupt the formation of the bar). Throughout this work,
we consider $t=0$ for the \Nbody{} run to be the time at which we added the
gas phase for the SMUGGLE run (i.e. we ignore the first $1.5\,\textrm{Gyr}$ of
evolution of the \Nbody{} disk when the bar is forming). We made one
additional modification when stitching the gas disk together with the \Nbody{}
disk - we created a hole within the central $4\,\textrm{kpc}$. This hole
guards against an initial dramatic infall of gas within the bar region, which
we found to destroy the bar. It is not uncommon for observed barrred galaxies
to have gas deficits in the bar region (though not in the very
center),\cite{1993RPPh...56..173S} and so our practice of allowing the gas
distribution to have a hole in the central region is consistent with our
choice to begin the simulations with a bar already formed.

\begin{figure}[t]%
\centering
\includegraphics[width=9cm]{fig/vcirc.pdf}
\caption{\textbf{The circular velocity curve of our initial setups.} This
curve is shown for the \Nbody{} run (blue) and the SMUGGLE run (orange)
compared to observational estimates for the Milky
Way.\cite{2019ApJ...871..120E} We see that the circular velocity curve for
both runs is marginally larger than the Milky Way's, but still comparable. The
SMUGGLE circular velocity curve is larger than the \Nbody{} curve due to the
additional mass in the gas phase.}
\label{fig:vcirc}
\end{figure}

>>>>

  Elongated bar-like features are ubiquitous in galaxies, occurring at the
  centers of approximately two-thirds of spiral
  disks\cite{2000AJ....119..536E, 2007ApJ...657..790M}.  Due to gravitational
  interactions between the bar and the other components of galaxies, it is
  expected that angular momentum and matter will redistribute over long (Gyr)
  timescales in barred galaxies \cite{1972MNRAS.157....1L,
  1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work ignoring the gas
  phase of galaxies has conclusively demonstrated that bars should slow their
  rotation over time due to their interaction with dark matter halos
  \cite{1992ApJ...400...80H, 2000ApJ...543..704D, 2002MNRAS.330...35A,
  2002ApJ...569L..83A, 2003MNRAS.341.1179A, 2003MNRAS.346..251O,
  2005MNRAS.363..991H, 2006ApJ...637..214M, 2007MNRAS.375..460W,
  2009ApJ...697..293D}. We have performed a simulation of a Milky Way-like
  galactic disk hosting a strong bar which includes a state-of-the-art model
  of the interstellar medium and a live dark matter halo. In this simulation
  the bar pattern does not slow down over time, and instead remains at a
  stable, constant rate of rotation. This behavior has been observed in
  previous simulations using more simplified models for the interstellar gas
  but it has remained unexplained\cite{1993AA...268...65F,
  2010ApJ...719.1470V}. We propose that the gas phase of the disk and the dark
  matter halo act in concert to stabilize the bar pattern speed and prevent the
  bar from slowing down or speeding up. We find that in a Milky Way-like disk,
  a gas fraction of only about \boldmath$5\%$ is necessary for this mechanism
  to operate. This result naturally explains why nearly all observed bars
  rotate rapidly\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
  2019MNRAS.482.1733G} and is especially relevant for our understanding of how
  the Milky Way arrived at its present state.

@@

\begin{figure*}[h!]%
\centering
\includegraphics[width=18cm]{fig/halo_wake.pdf}
\caption{\textbf{The wake excited in the dark matter halo.} The dark matter
halo wake is shown in the \Nbody{} case (\textit{middle panel}) and SMUGGLE case
(\textit{right panel}) after $2.6\,\textrm{Gyr}$ of evolution. The
\textit{middle} and \textit{right panels} show a surface density projection in
the $x$-$y$ plane of the dark matter halo after an axisymmetric average has
been subtracted. The solid line indicates the direction of the bar while the
dashed line indicates the direction of the halo wake (both measured by taking
the second Fourier component within a sphere of all material within a radius
of $4\,\textrm{kpc}$). The \textit{left panel} shows the time evolution of
the angle difference between the bar and the halo wake, as measured from the
second Fourier component. After the first Gyr, the angle difference in the
SMUGGLE case is smaller than in the \Nbody{} case by about a factor of two,
reflecting how the dark matter halo in the SMUGGLE case is unable to exert as
negative a torque on the bar as in the \Nbody{} case.}\label{fig:wake}
\end{figure*}

@@

It has long been known that non-axisymmetric features in a stellar disk act to
redistribute mass and angular momentum\cite{1972MNRAS.157....1L}. The
interaction between a bar and a spheroid (either a dark matter halo or stellar
bulge) has received considerable interest\cite{1984MNRAS.209..729T,
1985MNRAS.213..451W}. Stellar disks in isolation are prone to developing bars,
\cite{1971ApJ...168..343H} but the presence of a strong spherical potential
(e.g., from a stellar bulge or dark matter halo) acts to stabilize the disk
against bar formation\cite{1973ApJ...186..467O, 1976AJ.....81...30H}. Many
numerical simulations have confirmed that a live dark matter halo generically
exerts a negative torque on a bar, causing its pattern speed to decrease and
its length and strength to increase\cite{1992ApJ...400...80H,
2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. Intuitively,
the bar excites a wake of resonant material in a dark matter halo in a manner
akin to dynamical friction. This wake lags the bar and thus exerts a negative
torque on it. For a review of bar dynamics, see ref.\cite{1993RPPh...56..173S}

The expectation that a dark matter halo acts to slow down a bar is in
conflict with observational estimates of bar pattern speeds. Bar rotation
rates are typically classified by the dimensionless ratio
\begin{equation}
\Rot = \RCR/R_b\textrm{,}
\end{equation}
where \RCR{} is the radius of corotation and $R_b$ is the length of the
bar.\footnote{The radius of corotation \RCR{} is defined for circular orbits
as the radius at which the orbital frequency is equal to the pattern speed,
$\Omega_p$, of a given non-axisymmetric feature. In a galaxy with a constant
circular velocity $V_c$, it is given by $\RCR = V_c / \Omega_p$.} Galaxies
with $\Rot < 1.4$ are considered ``fast rotators'' while galaxies with $\Rot >
1.4$ are considered ``slow rotators''\cite{2000ApJ...543..704D}. Galaxies with
$\Rot < 1$ are not thought to be stable\cite{1980AA....81..198C}.
Observational estimates of the pattern speeds of bars indicate that nearly all
galaxies have $1 < \Rot < 1.4$\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G, 2020MNRAS.491.3655G}.

The role of gas on the evolution of the bar is less well-understood. Since the
gas phase typically only contributes about $10-20\%$ of the mass of a galaxy
at the present day, one might naively expect it to have a subdominant effect
on the bar. However, because gas is collisional, it can participate in
non-resonant angular momentum exchange with the bar\cite{2011MNRAS.415.1027H}.
Thus, numerical work has shown that the gas phase can have a stronger
influence on a bar than its contribution to the mass of a galaxy would
suggest\cite{2010ApJ...719.1470V, 2013MNRAS.429.1949A}.

We have performed a simulation of a disk galaxy using the finite-volume
gravito-hydrodynamics code \AREPO{}\cite{2010MNRAS.401..791S}. In \AREPO{},
the fluid is discretized as a moving Voronoi mesh. We use the additional
physics in the galaxy formation module Stars and MUltiphase Gas in GaLaxiEs
(SMUGGLE)\cite{2019MNRAS.489.4233M}. SMUGGLE is a comprehensive and
self-consistent galaxy formation model with a wide range of physical
processes, including radiative heating/cooling, star formation, and stellar
feedback. More detailed information on SMUGGLE is given in the Methods
section. Our disk galaxy is a modified version of the GALAKOS
model\cite{2020ApJ...890..117D}, which consists of a stellar disk, stellar
bulge, and dark matter halo. After $\sim2.5\,\textrm{Gyr}$ of evolution, the
GALAKOS disk forms a bar consistent with the Milky Way bar in terms of pattern
speed and length
($\sim40\,\textrm{km}/\textrm{s}/\textrm{kpc}$\cite{2022arXiv220606207G} and
$\sim4.5\,\textrm{kpc}$\cite{2015MNRAS.450.4050W}, respectively). We modify
this setup to include a gas phase, the details of which are provided in the
Methods section.

A mock stellar light image using filter profiles from the \textit{Hubble Space Telescope} is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the \Nbody{}
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large
qualitative difference in the evolution of the bar pattern between the two
runs. We see that in the \Nbody{} case, the bar lengthens in time and grows in
strength. In the SMUGGLE case, the bar retains a similar length and strength
between panels.

We show the time evolution of different bar properties in Fig.~\ref{fig:prop}.
In the upper panel, we show the pattern speed over time in the \Nbody{} (blue)
and SMUGGLE (orange) runs. The pattern speed in the \Nbody{} case slows down
while the pattern speed in the SMUGGLE case remains roughly constant. The
slowing down of the pattern speed in the \Nbody{} case is consistent with a long
line of numerical research on bars in \Nbody{}
simulations\cite{1992ApJ...400...80H, 2000ApJ...543..704D,
2002MNRAS.330...35A, 2002ApJ...569L..83A, 2003MNRAS.341.1179A,
2003MNRAS.346..251O, 2005MNRAS.363..991H, 2006ApJ...637..214M,
2007MNRAS.375..460W, 2009ApJ...697..293D}. However, in the SMUGGLE case the
pattern speed remains constant. After the first Gyr of evolution, we find that
the pattern speed increases by only $\sim10\%$ over the next
$4\,\textrm{Gyr}$, compared to a $\sim43\%$ decrease in the pattern speed for
the \Nbody{} run over the same interval. As we saw qualitatively in
Fig.~\ref{fig:overview}, the length of the bar in the \Nbody{} case grows over
time while it remains roughly constant in the SMUGGLE case. This is also
consistent with previous numerical work, which found that bars tend to grow as
they slow down and the radius of corotation
increases\cite{2000ApJ...543..704D, 2003MNRAS.341.1179A}.

@@

The fact that the dark matter halo in the SMUGGLE case exerts a smaller
negative torque on the bar can be understood in terms of the halo wake
mechanism. In the \Nbody{} case, halo material which is resonant with the bar
will form a wake which lags behind and exerts a negative torque on the bar,
slowing it down\cite{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}.\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its
pattern speed. However, the bar seems to empirically respond to a negative
torque induced by a halo wake by slowing down.} As the bar slows down, the
location of the resonances in phase space changes, allowing halo material
newly resonant with the bar to participate in forming the wake. However, the
gas is a reliable source of positive torque on the bar, speeding the bar up.
This stops the resonance location from changing such that the halo cannot
reinforce the wake, arresting the process by which the halo can slow the bar
down. We term this process ``exhaustion.''

@@

The presence of the gas can arrest the process by which additional material in
the dark matter halo can contribute to a wake. However, this does not explain
why the pattern speed in the SMUGGLE case is nearly constant over several Gyr.
Naively, it would be a coincidence that the bar pattern speed remains constant
in the SMUGGLE case, resulting from a chance cancellation of the halo and gas
torques. However, a constant pattern speed in the presence of gas has been
observed in a few simulations of barred galaxies with
gas\cite{1993AA...268...65F, 2010ApJ...719.1470V}. Previous work has argued
this is due to the bar torquing gas inwards, but no explanation has been given
for why it might remain constant.

We propose that an equilibrium mechanism is responsible for the pattern speed
remaining approximately constant. In this scenario, residual negative torque
from the dark matter halo balances out the positive torque from the gas phase.
It has been shown when an analytic bar is forced to rotate at a constant
pattern speed for a few Gyr, the halo exerts almost no torque on the
bar\cite{2022MNRAS.513..768C}. We instead see over a several Gyr time span
that the dark matter halo in our simulation is still able to support some
negative torque.

@@

We note that ref.\cite{2020MNRAS.491.3655G} found that the rotation parameter
\Rot{} positively correlates with gas fraction, such that galaxies with higher
gas fractions are rotating more slowly. However, it is not obvious this is in
tension with our result since the gas fraction of galaxies correlates with
other galactic properties\cite{2009ARAA..47..159B}. Furthermore, the
measurement of pattern speeds is a delicate process still prone to large error
bars.

The implications of our findings are numerous. First, we naturally explain why
nearly all observed galaxies are fast rotators without requiring the inner
regions of dark matter halos to be underdense\cite{1998ApJ...493L...5D,
2000ApJ...543..704D} or requiring new physics\cite{2021MNRAS.503.2833R,
2021MNRAS.508..926R}. Second, we show that the role of gas is of paramount
importance in studies which attempt to uncover the nature of dark matter from
its effect of slowing down the bar\cite{2021MNRAS.500.4710C,
2021MNRAS.505.2412C}. Third, we provide an explanation for how the Milky Way's
bar could be both long-lived and a fast rotator, of which there is some
observational evidence\cite{2019MNRAS.490.4740B}. And finally, we complicate
the picture of radial mixing expected to sculpt the Milky Way's disk
\cite{2012MNRAS.420..913B, 2015ApJ...808..132H}, a process which relies upon
the pattern speed of the bar to change with time (though our work does not
alter expectations for radial mixing induced by spiral
arms\cite{2002MNRAS.336..785S}).

Barred galaxies in cosmological simulations of galaxy formation continue to be
in conflict with observations by producing bars which rotate too
slowly\cite{2017MNRAS.469.1054A, 2019MNRAS.483.2721P, 2021AA...650L..16F}.
However, the pattern speeds of bars in both cosmological simulations and the
real universe can be affected by environmental processes not included in our
simulation -- e.g., satellite infall\cite{2011Natur.477..301P},
non-sphericity\cite{2013MNRAS.429.1949A} or rotation\cite{2013MNRAS.434.1287S,
2014ApJ...783L..18L, 2018MNRAS.476.1331C, 2019MNRAS.488.5788C} in the dark
matter halo, or perhaps even the gaseous circumgalactic medium. Naturally,
extending our present work to account for such effects is a crucial next step
in understanding the formation and evolution of galactic bars.

We found that below a certain gas fraction, bars should still be able to slow
down. Therefore, we expect barred spiral galaxies which have been gas-poor for
extended periods of time to be rotating very slowly. We therefore predict that
observations which target such galaxies (e.g., lenticular barred
galaxies\cite{2009ARAA..47..159B}) would find slowly rotating bars. There does
exist one example of a galaxy which is known to be a slow rotator -- the
low surface brightness galaxy UGC 628\cite{2009AA...499L..25C}. This galaxy
has been studied in detail by ref.\cite{2016MNRAS.463.1751C}, who note that it
indeed has a low gas fraction for galaxies of its type. We predict a general
trend that bars in gas-rich spiral galaxies should rotate quickly while bars
in gas-poor spiral galaxies should rotate slowly.

@@

\item[Data Availability]  Simulation output at a selection of timesteps as
well as the initial conditions will be made publicly available. Simulation
output at higher timestep frequency will be available upon request.

\item[Code Availability] A public version of the \AREPO{} code is available at
\url{https://arepo-code.org/}. An associated GitHub repository containing
analysis and plotting scripts will be made publicly available (currently
private). For the submission stage, we have included a tarball of this
repository (associated file \texttt{gasbar.tar.gz}).

\item[Supplementary Information] Supplementary information is available for
this manuscript.

@@

\noindent
{\bf Initial setup}
\\
\noindent
The initial setup of the galactic disk used in this work follows closely the
GALAKOS model\cite{2020ApJ...890..117D}, which uses a modified version of
the \texttt{MakeNewDisk} code\cite{2005MNRAS.361..776S}. The GALAKOS model has three
components - a radially exponential and vertically isothermal stellar disk,
and a stellar bulge and dark matter halo following a Hernquist
profile\cite{1990ApJ...356..359H}. All \Nbody{} runs in this work used the same
setup parameters as the GALAKOS disk, more details of which can be found in
the original paper.

The addition of the gas phase was done as follows. The version of
\texttt{MakeNewDisk} used for the original GALAKOS model can generate a gas
disk which is radially exponential and in vertical gravito-hydrodynamic
balance. We modified the radial profile of this code in order to allow us to
generate a disk with a constant surface density within some cut-off radius,
and then exponentially declining beyond that radius with the scale-length of
the stellar disk. Our fiducial model used an initial surface density of
$20\,M_{\odot}/\textrm{pc}^2$ and a cut-off radius of $9.3\,\textrm{kpc}$. The
initial gas disk is generated with a temperature of $10^4\,\textrm{K}$ and
solar metallicity.

After generating the gaseous disk in this way, we stitched the gas disk
together with the GALAKOS \Nbody{} disk (and bulge and dark matter halo) after
the GALAKOS disk has been allowed to evolve for $1.5\,\textrm{Gyr}$. The
purpose of allowing the GALAKOS disk to evolve first for a short period of
time is to allow for the bar to form unimpacted by the presence of the gas
(which would normally disrupt the formation of the bar). Throughout this work,
we consider $t=0$ for the \Nbody{} run to be the time at which we added the
gas phase for the SMUGGLE run (i.e. we ignore the first $1.5\,\textrm{Gyr}$ of
evolution of the \Nbody{} disk when the bar is forming). We made one
additional modification when stitching the gas disk together with the \Nbody{}
disk - we created a hole within the central $4\,\textrm{kpc}$. This hole
guards against an initial dramatic infall of gas within the bar region, which
we found to destroy the bar. It is not uncommon for observed barred galaxies
to have gas deficits in the bar region (though not in the very
center)\cite{1993RPPh...56..173S}, and so our practice of allowing the gas
distribution to have a hole in the central region is consistent with our
choice to begin the simulations with a bar already formed.

\begin{figure}[t]%
\centering
\includegraphics[width=9cm]{fig/vcirc.pdf}
\caption{\textbf{The circular velocity curve of our initial setups.} This
curve is shown for the \Nbody{} run (blue) and the SMUGGLE run (orange)
compared to observational estimates for the Milky
Way\cite{2019ApJ...871..120E}. We see that the circular velocity curve for
both runs is marginally larger than the Milky Way's, but still comparable. The
SMUGGLE circular velocity curve is larger than the \Nbody{} curve due to the
additional mass in the gas phase.}
\label{fig:vcirc}
\end{figure}

\begin{figure}[h!]%
\centering
\includegraphics[width=9cm]{fig/ps_GFM.pdf}
\caption{\textbf{Pattern speed evolution of a smooth ISM model.} This
evolution is shown for the fiducial disk in the \Nbody{} (blue), SMUGGLE
(orange), and smooth ISM (red) cases. The smooth ISM model is an older model
for the ISM which treats its multiphase nature in a subgrid
fashion\cite{2003MNRAS.339..289S}. This fundamentally differs from the SMUGGLE
model, which explicitly resolve the hot and cold phases of the
ISM\cite{2019MNRAS.489.4233M}. The pattern speed in the smooth ISM case is
broadly similar to the evolution in the SMUGGLE case. This shows that the
stability of the pattern speed is not simply a result of our assumed model for
the ISM.}
\label{fig:GFM}
\end{figure}
